Experiences Gamifying Developer Practices

### 2013-10-14 paper_id1.tex
Edited entire paper.  Worked out threats to validity section.  Added graphic with blaze tool image.

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.  We conducted this study in three stages, the pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivte better navigation practices and use of a code search tool.  We found that developers used the tool willingly and their feedback indicated the Blaze tool and game made them think about navigation practices more.  We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recomendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer`s work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancment.  As new techniques and tools are introduced to the development environment, developer may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.    For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.  Robillard et. al. show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.

@@

Gamification is the incorporation of game elements into an activity that people do not typcially consider a game\cite{2013Oxford}.  Why would software developers writing serious software be interested in this?  Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.

@@

During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We conducted post-pilot survey and feedback sessions that evaluated developers` acceptance of game elements in the tool and their desires for additional functionality.

This work contributes an assessment of developers` response to game-like elments in the promotion of new tools and practices.  We assess their early response through a survey of 130 software developers.  Then conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.  With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.

@@

Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.  Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.  This builds on this work by testing methods for increasing the use of structured navigation in developers` everyday practice.

@@

\subsection{Gamification}
Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

\section{Pre-study Sruvery on Viability of Applying Gamification for Software Engineering}

\subsection{Pre-study Background}
When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

@@

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}. It has some valuable lists of factors that motivate software engineers with the most common being ''the work'' motivates us. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges.

Understanding the concerns with maintaining intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the community would be receptive to a gamification approach to software development.   We designed  the survey to assess how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers, and being recognized with badges and points. The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.

@@

Using a question from the Saachi and Saachi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?  We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  The overall response from our survey is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.  
%EMH why is our survey result comparable to Saatchi?
\begin{figure}
	\includegraphics[width=3.4in]{gamificationquestion.pdf}
	\caption{Acceptance of gamification}
	\label{fig:gamification}
\end{figure}

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable. Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.  
Two conclusions from this feedback are that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

Developers were asked to rank awards for different actions by which they would find the most desirable.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.
%EMH connection?
\begin{itemize}
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current work-flow pattern the developer follows.
\item Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.  So, the tool will incorporate team competition ideas and recognize quality practices.
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}

\section{Game Design}

@@

%distinguish pilot from pre-study.  How does gamification apply?
The pilot evaluates the aspects of game implementation as well as the study of the results of their results. Here we discuss  relevant aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.

\subsection{Select a Practice}
%transition
In their paper on game design patterns, Hamari et. al. issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .  This warning indicates that we should avoid giving points for things that are already an expected deliverable such as finding or fixing bugs, completing development tasks, or meeting required process criteria.  
%passive
Practices to consider are those practices with room to go beyond a checklist or minimum criteria where developers have the opportunity to excel.
%passive
Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.  We currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.  
%EMH is check-in frequency a practice?
Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.  Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro thus may provide a candidate for future gamification application.   
%reference
Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.

Based on studies of structured navigation by Robillard et. al. that identify structured navigation as a practice leading to effective program maintenance  \cite{wbsnipes:Robillard2004How}, we identified structured navigation as a key practice.   Structured navigation fits the requirement of a practice that is contained in the IDE, is something that has proven benefits, and lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.

@@

We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.  For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code, which we classify as structured navigation.

@@

To evaluate developers using recommended tools in Visual Studio, we categorized them by their tool name.  In the pilot we were interested in developers learning about and using the Sando search tool.  Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.

@@

\subsection{Competitive Elements}
By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.    The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.  However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.  Therefore the assigned letters could even change when an individual's ranking changed.  Not knowing who of your colleagues was ahead of you could be a factor in motivation, but also avoids one potential source of conflict in an intact team.

@@

In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.  The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.  This final configuration continued until the end of the pilot.

@@

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.  Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
%Through this approach, we ensured that the developers are performing the task on a very natural environment with out bothering that they are being tracked.
\subsection{Pilot Survey}

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers` navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

The key questions about this pilot were did we have an effect on knowledge of the developers regarding structured navigation commands.  We asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.  We asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.

The other key question is did we have an effect on their navigation practices.  Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of not at all, some, many times, and as much as possible.  Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.  These two questions assessed their intent in terms of using structured navigation commands and we checked that with the data.

A key set of questions focused on whether using Blaze increased their attention to navigation commands by asking how much specific game elements influenced their practices.  The scale started with not at all, a little, some, to a lot.

@@

@@

%\subsubsection{Observations About Developers` Habits}
%
%We are interested in understanding whether developers are consistently using practices that make them inefficient.  One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.  If this is possible, then we can discuss whether developer have consistent ways of navigating and understanding for each edit.  Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.
%
%To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.  Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.  The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.  
%
%While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.  Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\

@@

We selected navigation as the subject of the pilot game because it was identified by Robillard et. al\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    Are we focusing on the right area of developers` activity in Visual Studio?

@@

This analysis shows that we cannot disprove that unstructured navigation is correlated with edit duration, thus it is an important factor in developer productivity.  Combining unstructured navigation with reading describes a larger portion of the overall edit duration and motivates us towards making tools that improve both navigation methods and code comprehension.

@@

What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.  We evaluated the week over week differences using a Wilcoxon Rank sum test that compares week one to week two and week one to week three.    In week two when the developers were able to learn about structured navigation commands there was no significant change.  Also in week two, two developers stopped participating, one was working in another environment and the other dropped of for the rest of the study.  Week three there was a slight increase, however the increase is not significant.  Weeks four and five that follow the formal period of study show the navigation ratio continues in a similar range.  Overall, we did not see a significant effect towards increasing navigation ratio for the interventions Blaze provided to the environment.

@@

The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.  Per the pilot design, the points display was disabled until the third week following installation so that a baseline without instant feedback could be established.  The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features.  Observing the result shows that during the first two weeks, we see in Figure \ref{fig:pointsbyweek} developers were fairly consistent in their use patterns.  With the introduction of the points feedback, two developers` points accumulation spiked in week three due mostly to increased use of Sando.   We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.

@@

We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing. Thus categories like testing, debug, and reading were affected by whether or not they used the virtual environment or their desktop for these activities.  Individual differences in those categories are considered irrelevant and not discussed here. Because our game points analysis focused on navigation commands excluding reading, the presence of the virtual environment may explain how the data have missing values on some days.

We also discussed with one developer who`s build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.  We asked the developer to explain how building more frequently helps them. The developer mentioned that by doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far. The developer feels that if (s)he resolves the bugs then and there, (s)he can do it faster as the code and logic is fresh in her memory. If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it. Hence the developer feels that more frequent builds improves their productivity.

@@

%Though commonly held opinion of Microsoft`s Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices. 
Developers also reiterated that they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.  Moving towards more immediate feedback, we shared a set of achievement badges tied to specific navigation commands  as we gave the more quantitative feedback to the developers.   Developers said that badges would motivate them to use some specific tools or commands more than points alone.   This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.  The important factor in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.

We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers` perception of their practices and their demographic information.

>>>>

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.  We conducted this study in three stages, the pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivate better navigation practices and use of a code search tool.  We found that developers used the tool willingly and their feedback indicated the Blaze tool and game made them think about navigation practices more.  We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancement.  As new techniques and tools are introduced to the development environment, developer may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.    For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.  Robillard et. al. show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.

@@

Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.  Why would software developers writing serious software be interested in this?  Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.

@@

During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.

This work contributes an assessment of developers' response to game-like elements in the promotion of new tools and practices.  We assess their early response through a survey of 130 software developers.  Then conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.  With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.

@@

Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.  Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.  This builds on this work by testing methods for increasing the use of structured navigation in developers' everyday practice.

@@

\subsection{Gamification}

When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

@@

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}.  They provide a useful inventory of aspects of the job motivate software engineers with the most common being ''the work'' motivates us. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges. 
Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

\section{Pre-study Sruvery}

\subsection{Pre-study Background}

Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.   We designed  the survey to assess how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers, and being recognized with badges and points. The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?  We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% of 18-45 year olds that Saatchi and Saatchi found were at least somewhat interested.

\begin{figure}
	\includegraphics[width=3.4in]{gamificationquestion.pdf}
	\caption{Acceptance of gamification}
	\label{fig:gamification}
\end{figure}

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  A graduated scale was used for the scope of the shared community and two questions distinguished between whether that sharing was anonymous or the information was personally identifiable. Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.  
Two conclusions from this feedback are that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

Developers were asked to rank awards for different actions by which they would find the most desirable.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  This supports a conclusion that focusing on awarding quality practices would motivate developers to do them consistently.

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers favor a team-oriented competition competition over individual awards
\item Awards for performing good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool that focused on sharing and rewarding developers for using good practices and tools. 
\section{Game Design}

@@

%distinguish pilot from pre-study.  How does gamification apply?

Following the pre-study results, we began implement a game-like tool for developers to use in Visual Studio.  We defined the goals for the tool and designed the game elements that it would include.  Here we discuss  relevant aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.

\subsection{Select a Practice}
The first step is to select a development practice to incorporate into the game.  
Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.  In their paper on game design patterns, Hamari et. al. issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .  Thus we avoided giving points for actions like fixing bugs or completing tasks.  We also currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.  
%EMH is check-in frequency a practice?

Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.  Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro \cite{V:Johnson2007Automated} thus may provide a candidate for future gamification application.   
%reference for debugging?
Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.

Studies of developer navigation by Robillard et. al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice that more effective developers use when maintaining programs.   Structured navigation fits the requirements of a practice that is contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.  Thus we selected structured navigation as the practice to demonstrate gameification.

@@

We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.  These commands include Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} that are built-in to Visual Studio and follow the structured graph of the code.

@@

To evaluate developers using recommended tools in Visual Studio, we categorized them by their tool name.  In the pilot we wanted developers to learn about and adopt the Sando search tool.  Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.

@@

\subsection{Competitive Elements}
By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.    The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.  However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.  Therefore the assigned letters could even change when an individual's ranking changed.  Not knowing who of your colleagues was ahead of you could be a factor that reduced motivation, but also avoids one potential source of conflict in an intact team.

@@

In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The Info button provides a level indicator via colored contents and activates a chart display windows when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderbaord tab to see how their scores compared with the others in the pilot.  This final configuration remained as the display until the end of the pilot.

\begin{figure}
	\centering
	\includegraphics[width=2.2in]{blazeWindow.png}
	\caption{Blaze Tool Window}
	\label{fig:blazeWindow}
\end{figure}

@@

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.  Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.

\subsection{Pilot Survey}

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

To find out whether Blaze had an effect on knowledge of the developers regarding structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.  Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.

The other key question is did Blaze have a perceived effect on their navigation practices.  Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of not at all, some, many times, and as much as possible.  Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.  These two questions assessed their perceptions of their structured navigation command use that we compared with the data collected from Blaze.

The final assessment questions tested whether using Blaze influenced their attention to navigation commands by asking how much specific game elements influenced their practices.  The scale started with not at all, a little, some, to a lot.

@@

We conducted an in-person post pilot survey in order to cross check the quantitative data from Blaze with developers' perception of their practices and their demographic information.

The pilot survey queried developers' relevant demographic data to help understand some differences in results.  We asked developers to provide their years of experience so that we could evaluate any effects experience might have.  The developers in the study had between 7 and 14 years experience.  One developer who dropped out of the study said that their lower level of experience compared to their peers influenced their decision to stop participating.

On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.  Other half of the participants have some understanding on structured navigation use it some of the time. Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their colleagues.  While analyzing the points they scored with Blaze tool, we observed that the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.

The participants requested that Blaze provide more feedback on tips to help them increase their score. They indicated that more obvious feedback such as hints that pop-up when they launch the tool would help them more than passively provided information.  Their desire for personal metrics and historical views of the data encourages us that this detailed level of measurement and analysis is perceived as helpful.  Features to meet these requests are part of future work.

@@

%\subsubsection{Observations About Developers' Habits}
%
%We are interested in understanding whether developers are consistently using practices that make them inefficient.  One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.  If this is possible, then we can discuss whether developer have consistent ways of navigating and understanding for each edit.  Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.
%
%To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.  Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.  The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.  
%
%While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.  Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\

@@

We selected navigation as the subject of the pilot game because it was identified by Robillard et. al\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis we ask whether we are focusing on the right area of developers' activity in Visual Studio?

@@

This analysis shows that we cannot disprove that unstructured navigation is correlated with edit duration, thus we conclude that it is an important factor in developer productivity.

@@

What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.  We evaluated the week over week differences using a Wilcoxon Rank sum test \cite{RefWorks:118} that compares week one to week two and week one to week three.    In week two when the developers were able to learn about structured navigation commands there was no significant change.  Also in week two, two developers stopped participating in the study.  One was working in another environment and returned, while the other dropped for the rest of the study.  Week three there was a slight increase in the mean, however the increase is not significant.  Weeks four and five that follow the formal period of study show the navigation ratio continues in a similar range.  Overall, we did not see a significant effect towards increasing navigation ratio for the interventions Blaze provided to the environment.

@@

The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.  Per the pilot design, the points display was disabled until the third week following installation so that a baseline without instant feedback could be established.  The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features. The result shows that during the first two weeks shown in Figure \ref{fig:pointsbyweek}, developers were fairly consistent in their use patterns.  With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.   We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.  In these results we see an initial though not significant effect from the points display in Blaze that is not sustained over the remaining weeks.  We found following discussions that developers wanted more guidance and more prompting included in the game which we discuss further in the next section.

@@

We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing. Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.  Individual differences in those categories are considered irrelevant and not discussed here. The presence of the virtual environment may explain how some developers have no activity recorded on some days.

%Can we get a quote here?
One developer's build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.  We asked the developer to explain how building more frequently helps them. The developer reported that by doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far. The developer feels that (s)he resolves the bugs faster with frequent builds because the code and logic is fresh in her memory. If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it. Hence the developer feels that more frequent builds improves their productivity.

@@

%Though commonly held opinion of Microsoft's Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices. 
Developers also reiterated that they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.  Moving towards more immediate feedback, we shared a set of achievement badges tied to specific navigation commands  as we gave the more quantitative feedback to the developers.   Developers said that badges would motivate them to use some specific tools or commands more than points alone. 
%  This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.  The important difference in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.

\section{Threats to Validity}
The threats to construct validity occur when measurements may not reflect the intended operational definition.The key measures in our study that could have results differing from their definition were navigation ratio and points scored in the game.

Measuring navigation ratio involved categorizing events into structured and unstructured navigation.  The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structure navigation depending on how the developer uses some commands like Find.  Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, we might consider that something other than navigation (though it is classified as such for this paper).
